---
title: "R Notebook"
output: html_notebook
---


Hashing Exercise

1. Use md5, sha1 and sha256 hashing algorithms to anonymize the following text:

  a. Adaire Academy
  b. adaire academy
  c. ADAIRE ACADEMY
  d. adaireacademy
  
What do you notice?

2. Hash the following numeric vector : c(1, 2, 4, 6).
What is your result?

3. Convert the vector in 2 above to a characater string before hashing it. Is there a difference?
  a. Try the integer: 12345 and the character "12345". Is the hashcode the same?

4. Create a txt file called 'example.txt' with some simple text 'My name is Chantelle', and hash the file, the modify the content of the file slightly and hash it again. Is there a noticeable difference?
Steps
  a. Create a txt file in r using the following code structure:
  writelines("My name is Chantelle", "example.txt")
  digest(file = "example.txt", algo="md5")
  
Now modify the content of the file, adding a punctuation to the end of the content of your txt file and run the hash algorithm. Is it different?

5. Create a function called password_checker, that takes in a user's password, and checks whether it's hash is the same has a stored hash code for the password.
return "sucessful" if it is the same and "unsuccessful, try again" if it is not.

function name: password_checker
function inputs: passwords, stored_hashes (a dataframe containing passwords and their hashes)
function output: "sucessful" or "unsuccessful, try again".

# Load the digest package
library(digest)

# Sample data: usernames and their passwords
usernames <- c("chantelle", "alex", "maria", "daniel")
passwords <- c("sunshine123", "helloWorld", "qwerty456", "moonlight!")

# Hash the passwords using SHA-256
password_hashes <- sapply(passwords, function(pw) digest(pw, algo = "sha256"))

# Create the datagram (data frame)
password_db <- data.frame(
  username = usernames,
  password = passwords,
  hash = password_hashes,
  stringsAsFactors = FALSE
)

```{r}
# Load the digest package
library(digest)

# Sample data: usernames and their passwords
usernames <- c("chantelle", "alex", "maria", "daniel")
passwords <- c("sunshine123", "helloWorld", "qwerty456", "moonlight!")

# Hash the passwords using SHA-256
password_hashes <- sapply(passwords, function(pw) digest(pw, algo = "sha256"))

# Create the datagram (data frame)
password_db <- data.frame(
  username = usernames,
  password = passwords,
  hash = password_hashes,
  stringsAsFactors = FALSE
)

password_db
```
```{r}
password_checker <- function(password, db){
  password_col <- db$passwords
  password_hash <- db$hash
  pwd_hash <- digest(password, algo="sha256")
  for (i in password_col){
    if (password %in% password_col){
      if (pass_word)
    }
  }
}
```





6. 
  
