---
title: "R Notebook"
output: html_notebook
---

#  STEP 1: Install and Load the Image Package
# This package lets us load and work with images like PNG, JPG, BMP, etc.

```{r}
install.packages("imager")  # Only run this the first time
library(imager)
```

#  STEP 2: Load Your Image
# Replace "your_image.png" with the path to your image file.
# Think of this as opening a picture so we can work on it.

```{r}
img <- load.image("sample_image.png")
plot(img)
```

```{r}
dim(img)
```

# Optional: Resize if image is too large (makes it faster to work with)

```{r}
img <- resize(img, -50, -50)  # Reduce size by 50% in both directions
plot(img1)
dim(img1)
```

# Show the original image

```{r}
plot(img, main = "Original Image")
```


#  STEP 3: Get the Image Data
# We now convert the image into numbers.
# Every picture on your screen is just made of thousands of tiny squares (pixels).
# Each pixel has a Red, Green, and Blue value that tells the screen how to light up.

```{r}
img_array <- as.array(img)
dim(img_array)  # Shows the shape of the image [width, height, color, frame]
```

```{r}
img_array
```


#  STEP 4: Split the Image into Colors
# We break the image into Red, Green, and Blue parts so we can compress each separately.

```{r}
red   <- img_array[,,1,1]  # Red channel
green <- img_array[,,1,2]  # Green channel
blue  <- img_array[,,1,3]  # Blue channel
```

#  STEP 5: Create a Function to Compress Each Color
# Here's the trick: Instead of storing all the tiny details of the picture,
# we'll keep only the most important parts using a method called SVD.
# You can think of this like summarizing a big book in a few key points!

```{r}
compress_color <- function(channel_matrix, keep_top_k = 50) {
  # SVD breaks the image into parts we can choose from.
  svd_result <- svd(channel_matrix)
  
  # We only keep the first 'k' most important parts
  u <- svd_result$u[, 1:keep_top_k]
  d <- diag(svd_result$d[ 1:keep_top_k ])
  v <- svd_result$v[, 1:keep_top_k]
  
  # This puts the image back together using only the important parts.
  compressed <- u %*% d %*% t(v)
  
  return(compressed)
}
```


#  STEP 6: Compress Each Color Channel
# Try changing the value of `k` below to see how much detail you want to keep.

```{r}
k <- 7  # More = better quality, less = smaller file size

red_compressed   <- compress_color(red, k)
green_compressed <- compress_color(green, k)
blue_compressed  <- compress_color(blue, k)
```


#  STEP 7: Put the Colors Back Together
# Now we combine the compressed Red, Green, and Blue parts back into one image.
# Also make sure the color values stay between 0 and 1 (that's what screens use).

```{r}
compress_clip <- function(x) {
  x[x < 0] <- 0
  x[x > 1] <- 1
  return(x)
}

reconstructed <- array(0, dim = dim(img_array))
reconstructed[,,1,1] <- compress_cli p(red_compressed)
reconstructed[,,1,2] <- compress_clip(green_compressed)
reconstructed[,,1,3] <- compress_clip(blue_compressed)

compressed_img <- as.cimg(reconstructed)
```


#  STEP 8: Show the Compressed Image
# You’ll notice the image still looks almost the same, but it uses much fewer data!

```{r}
par(mfrow = c(1, 2))  # Show side-by-side
plot(img, main = "Original Image")
plot(compressed_img, main = paste("Compressed Image (k =", k, ")"))
```

#  STEP 9: Save the Compressed Image (Optional)
# You can save the compressed version as a new file.

```{r}
save.image(compressed_img, "compressed_output.png")
```





