---
title: "R Notebook"
output: html_notebook
---

#  STEP 1: Install and Load the Image Package
# This package lets us load and work with images like PNG, JPG, BMP, etc.

```{r}
# install.packages("imager")  # Only run this the first time
library(imager)
```

#  STEP 2: Load Your Image
# Replace "your_image.png" with the path to your image file.
# Think of this as opening a picture so we can work on it.

```{r}
img <- load.image("sample_image.png")
plot(img)
```

```{r}
dim(img)
```

```{r}
# for the original image, the value of k is the smaller dimension among x and y axes of the image.
(dim(img)[1] * dim(img)[2])/ (dim(img)[2] * 3)
```

# Optional: Resize if image is too large (makes it faster to work with)

```{r}
img <- resize(img, -50, -50)  # Reduce size by 50% in both directions
plot(img1)
dim(img1)
```

# Show the original image

```{r}
plot(img, main = "Original Image")
```


#  STEP 3: Get the Image Data
# We now convert the image into numbers.
# Every picture on your screen is just made of thousands of tiny squares (pixels).
# Each pixel has a Red, Green, and Blue value that tells the screen how to light up.

```{r}
img_array <- as.array(img)
dim(img_array)  # Shows the shape of the image [width, height, color, frame]
```

```{r}
img_array
```


#  STEP 4: Split the Image into Colors
# We break the image into Red, Green, and Blue parts so we can compress each separately.

```{r}
red   <- img_array[,,1,1]  # Red channel
green <- img_array[,,1,2]  # Green channel
blue  <- img_array[,,1,3]  # Blue channel
```

#  STEP 5: Create a Function to Compress Each Color
# Here's the trick: Instead of storing all the tiny details of the picture,
# we'll keep only the most important parts using a method called SVD.
# You can think of this like summarizing a big book in a few key points!

```{r}
compress_color <- function(channel_matrix, keep_top_k = 50) {
  # SVD breaks the image into parts we can choose from.
  svd_result <- svd(channel_matrix)
  
  # We only keep the first 'k' most important parts
  u <- svd_result$u[, 1:keep_top_k]
  d <- diag(svd_result$d[1:keep_top_k])
  v <- svd_result$v[, 1:keep_top_k]
  
  # This puts the image back together using only the important parts.

  compressed <- u %*% d %*% t(v)
  
  return(compressed)
}
```


#  STEP 6: Compress Each Color Channel
# Try changing the value of `k` below to see how much detail you want to keep.

```{r}
k <- 7  # More = better quality, less = smaller file size

red_compressed   <- compress_color(red, k)
green_compressed <- compress_color(green, k)
blue_compressed  <- compress_color(blue, k)
```


#  STEP 7: Put the Colors Back Together
# Now we combine the compressed Red, Green, and Blue parts back into one image.
# Also make sure the color values stay between 0 and 1 (that's what screens use).

```{r}
compress_clip <- function(x) {
  x[x < 0] <- 0
  x[x > 1] <- 1
  return(x)
}

reconstructed <- array(0, dim = dim(img_array))
reconstructed[,,1,1] <- compress_cli p(red_compressed)
reconstructed[,,1,2] <- compress_clip(green_compressed)
reconstructed[,,1,3] <- compress_clip(blue_compressed)

compressed_img <- as.cimg(reconstructed)
```


#  STEP 8: Show the Compressed Image
# You’ll notice the image still looks almost the same, but it uses much fewer data!

```{r}
par(mfrow = c(1, 2))  # Show side-by-side
plot(img, main = "Original Image")
plot(compressed_img, main = paste("Compressed Image (k =", k, ")"))
```

#  STEP 9: Save the Compressed Image (Optional)
# You can save the compressed version as a new file.

```{r}
save.image(compressed_img, "compressed_output.png")
```

# Practical Exercises


### Question 1

Modify the provided code to:
1) Test three different k-values: 5, 25, and 50 on the same image. 
2) Display all three compressed versions side by side in a single plot. 
3) Which k-value provides the best balance between file size and quality for your image?

```{r}
img <- load.image("sample_image.png")
img_array <- as.array(img)
compressed_image_names <- c("compressed_k=5.png", "compressed_k=25.png", "compressed_k=30.png", "compressed_k=50.png", "compressed_k=75.png")

red   <- img_array[,,1,1]  # Red channel
green <- img_array[,,1,2]  # Green channel
blue  <- img_array[,,1,3]  # Blue channel

k_values <- c(5, 25, 30, 50, 75)
n <- length(k_values) + 1
par(mfrow=c(2, n/2))
plot(img, main="Original Image", axes=FALSE)


for (i in 1:length(k_values)) {
      k <- k_values[i]
      red_compressed   <- compress_color(red, k)
      green_compressed <- compress_color(green, k)
      blue_compressed  <- compress_color(blue, k)
      
      reconstructed <- array(0, dim = dim(img_array))
      reconstructed[,,1,1] <- compress_clip(red_compressed)
      reconstructed[,,1,2] <- compress_clip(green_compressed)
      reconstructed[,,1,3] <- compress_clip(blue_compressed)
      
      compressed_img <- as.cimg(reconstructed)
      name <- compressed_image_names[i]
      save.image(compressed_img, name)
      plot(compressed_img, main = paste("Compressed Image (k =", k, ")"), axes=FALSE)
      }

```

```{r}
library(stringr)
for (name in compressed_image_names){
  print(paste("The size of image with k =", str_extract(name, "\\d+"), "is: ", as.character(round(file.info(name)$size / 1024), 2), "megabytes"))
}

print(paste("The size of the original image is: ", as.character(round(file.info("sample_image.png")$size/1024), 2), "megabytes"))


```
The k value that gives a god balance between size and quality usually depends on the usecase, but generally is 25 or 30  for this image.


### Question 2

The original code uses k=7. Change this value to k=1 and run the compression. What happens to the image? Now try k=100. Describe the differences you observe and explain why this occurs.

```{r}
img <- load.image("sample_image.png")
img_array <- as.array(img)
compressed_image_names <- c("compressed_k=2.png", "compressed_k=100.png")

red   <- img_array[,,1,1]  # Red channel
green <- img_array[,,1,2]  # Green channel
blue  <- img_array[,,1,3]  # Blue channel

k_values <- c(2, 100)
n <- length(k_values)
par(mfrow=c(1, 2))
#plot(img, main="Original Image", axes=FALSE)


for (i in 1:length(k_values)) {
      k <- k_values[i]
      red_compressed   <- compress_color(red, k)
      green_compressed <- compress_color(green, k)
      blue_compressed  <- compress_color(blue, k)
      
      reconstructed <- array(0, dim = dim(img_array))
      reconstructed[,,1,1] <- compress_clip(red_compressed)
      reconstructed[,,1,2] <- compress_clip(green_compressed)
      reconstructed[,,1,3] <- compress_clip(blue_compressed)
      
      compressed_img <- as.cimg(reconstructed)
      name <- compressed_image_names[i]
      save.image(compressed_img, name)
      plot(compressed_img, main = paste("Compressed Image (k =", k, ")"), axes=FALSE)
      }
```
The more the k the more the significant values and thus there more details are retained in the image.



### Question 3

Add code to calculate and display the compression ratio for each k-value you test. The compression ratio can be estimated as: original_dimensions / (k * 3). Which k-value gives you approximately 90% compression?

```{r}
img <- load.image("sample_image.png")
img_array <- as.array(img)
compressed_image_names <- c("compressed_k=5.png", "compressed_k=25.png", "compressed_k=30.png", "compressed_k=50.png", "compressed_k=75.png")

red   <- img_array[,,1,1]  # Red channel
green <- img_array[,,1,2]  # Green channel
blue  <- img_array[,,1,3]  # Blue channel

k_values <- c(5, 25, 30, 50, 75)
n <- length(k_values) + 1
par(mfrow=c(2, n/2))
plot(img, main="Original Image", axes=FALSE)


for (i in 1:length(k_values)) {
      k <- k_values[i]
      red_compressed   <- compress_color(red, k)
      green_compressed <- compress_color(green, k)
      blue_compressed  <- compress_color(blue, k)
      
      reconstructed <- array(0, dim = dim(img_array))
      reconstructed[,,1,1] <- compress_clip(red_compressed)
      reconstructed[,,1,2] <- compress_clip(green_compressed)
      reconstructed[,,1,3] <- compress_clip(blue_compressed)
      
      compressed_img <- as.cimg(reconstructed)
      image_dimension <- dim(img)
      compression_ratio <- (image_dimension[1] * image_dimension[2]) / ( (image_dimension[2] + image_dimension[1] + 1)* k) 
      name <- compressed_image_names[i]
      save.image(compressed_img, name)
      print(paste("compression ratio for k ="  ,k, "is", round(compression_ratio, 2)))
      plot(compressed_img, main = paste("Compressed Image (k =", k, "),", compression_ratio, "%"), axes=FALSE)
      }

```


```{r}
# for the original image, the value of k is the smaller dimension among x and y axes of the image.
# compression ratio of the original
(dim(img)[1] * dim(img)[2])/ (dim(img)[2] * (dim(img)[1] + dim(img)[2] + 1))
```
None, but k = 5 gives 70 which is the closest to 90% compression.

### Question 4

Modify the code to compress only the red channel while leaving green and blue channels uncompressed.Display the result alongside the original image. What visual effects do you notice?

```{r}
img <- load.image("sample_image.png")
img_array <- as.array(img)
compressed_image_names <- c("compressed_k=5.png", "compressed_k=25.png", "compressed_k=30.png", "compressed_k=50.png", "compressed_k=75.png")

red   <- img_array[,,1,1]  # Red channel
green <- img_array[,,1,2]  # Green channel
blue  <- img_array[,,1,3]  # Blue channel

k_values <- c(25, 30, 50)
n <- length(k_values) + 1
par(mfrow=c(2, n/2))
plot(img, main="Original Image", axes=FALSE)


for (i in 1:length(k_values)) {
      k <- k_values[i]
      red_compressed   <- compress_color(red, k)
      green_compressed <- compress_color(green, k)
      blue_compressed  <- compress_color(blue, k)
      
      reconstructed <- array(0, dim = dim(img_array))
      reconstructed[,,1,1] <- compress_clip(red_compressed)
      reconstructed[,,1,2] <- compress_clip(green)
      reconstructed[,,1,3] <- compress_clip(blue)
      
      compressed_img <- as.cimg(reconstructed)
      image_dimension <- dim(img)
      compression_ratio <- (image_dimension[1] * image_dimension[2]) / ( (image_dimension[2] + image_dimension[1] + 1)* k) 
      name <- compressed_image_names[i]
      save.image(compressed_img, name)
      print(paste("compression ratio for k ="  ,k, "is", round(compression_ratio, 2)))
      plot(compressed_img, main = paste("Compressed Image (k =", k, "),", compression_ratio, "%"), axes=FALSE)
      }

```




