---
title: "Learning SQL-like Queries in R with {dplyr}"
author: "Chantelle"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Theme 1: Connecting R to Databases

### Objective
Learn to connect to a SQLite database using R and explore tables using `DBI` and `RSQLite`.

### Packages Required
```{r}
install.packages("DBI")
install.packages("RSQLite")
install.packages("dplyr")
install.packages("tibble")
```

### Step-by-Step Connection
```{r}
library(DBI)
library(RSQLite)
library(dplyr)
library(tibble)

# Set up the database connection
shop_db <- dbConnect(SQLite(), "shop.db")

# List all tables in the database
dbListTables(shop_db)
```

### Note
Make sure `shop.db` is in your R working directory. Use `getwd()` to confirm. You can move to it using `setwd("your/folder/path")` if needed.

### Inspecting Tables
```{r}
# Connect to customers table
customers_tbl <- tbl(shop_db, "customers")

# Display a preview
customers_tbl
```

### Understanding What You See
You’re seeing a preview because `{dplyr}` delays full execution. This is called lazy evaluation and it avoids loading all data into memory.

### Exercise 1: Explore Table Columns
Use `select()` to show only the `first_name` and `age` columns.

```{r}
# Your code here
customers_tbl %>% 
  select(first_name, age)
```

### Exercise 2: Filter Rows
Show only customers older than 25.

```{r}
customers_tbl %>% filter(age > 25)
# Your code here
```

### Exercise 3: Add New Columns
Create a new column `age_in_5` which is age + 5.

```{r}
customers_tbl %>% mutate(age_in_data = age + 5)
# Your code here
```

### Notes on Lazy Evaluation
- `tbl()` creates a *lazy* reference to the table. It doesn't load the full data into memory.
- To execute the pipeline and retrieve data, use `collect()`.

### Exercise 4: Use `collect()`
Collect the results of your previous query.

```{r}
customers_tbl %>% mutate(age_in_data = age + 5) %>% collect()
# Your code here
```

### Exercise 5: Load and explore the `sales` table
```{r}
sales_tbl <- tbl(shop_db, "sales")
sales_tbl
```

### Exercise 6: Count total number of rows in `sales_tbl`
> Hint: use `count()` on a constant column like `customer_id` or just use `tally()`.

```{r}
# Your code here
sales_tbl %>% select(customer_id) %>% count()
sales_tbl %>% tally()
```

### Exercise 7: Which customer had the highest number of purchases?
Group by `customer_id`, count, then join with `customers_tbl` to get name.

```{r}
sales_tbl %>% collect()
```

```{r}
sales_tbl %>% group_by(customer_id) %>% tally(sort=T) %>%  collect()
```


```{r}
sales_tbl %>% 
  group_by(customer_id) %>% 
  tally(sort=TRUE) %>% 
  left_join(customers_tbl, by = 'customer_id') %>% 
  select(customer_id, first_name, last_name, n) %>% 
  collect()
```

### Bonus Tip
Try writing your code in blocks and running each one. Don’t be afraid to print intermediate results to understand what’s happening.

## Theme 2: Replacing SQL with `{dplyr}`

### Objective
Use `{dplyr}` to perform common SQL-like tasks without writing SQL.

### Intuition
The functions in `{dplyr}` closely map to SQL keywords:

| dplyr Function | SQL Equivalent  |
|----------------|------------------|
| select()       | SELECT           |
| filter()       | WHERE            |
| arrange()      | ORDER BY         |
| mutate()       | column creation  |
| group_by() + summarise() | GROUP BY + aggregation |
| left_join()    | LEFT JOIN        |

### Exercise 8: SQL to dplyr Translation
**SQL:**
```sql
SELECT first_name FROM customers WHERE age > 30;
```
**dplyr:**
```{r}
# Your code here
customers_tbl %>% filter(age > 30) %>% select(first_name) %>% collect()
```

### Exercise 9: Multiple Filters
Show `first_name`s of customers who are male or under 18, sorted by age.

```{r}
# Your code here
customers_tbl %>% filter(gender == 'Male' | age < 18 ) %>% arrange(age) %>% collect
```

### Exercise 10: Count by Group
Count how many customers exist by gender.

```{r}
# Your code here
customers_tbl %>% group_by(gender) %>% tally()
```

### Exercise 11: Join Tables
Join `customers_tbl` with `sales_tbl` on `customer_id`.

```{r}
# Your join code here
customers_tbl %>% left_join(sales_tbl, by='customer_id')
```

### Exercise 12: Add Combined Name
Create a new column `name` combining `first_name` and `last_name` using `paste()`.

```{r}
# Your code here
customers_tbl %>% mutate(name = paste(first_name, " ", last_name))
```

### Exercise 13: Summary Aggregation
Get average price of sales by gender. Use the joined table.

```{r}
# Your code here
customers_tbl %>% left_join(sales_tbl, by = "customer_id") %>% select(gender, price) %>% group_by(gender) %>% summarise(avg_price = mean(price))
```

### Exercise 14: Using `show_query()`
Use `show_query()` on your joined pipeline from above. What SQL is generated?

```{r}
# Your code here
customers_tbl %>% left_join(sales_tbl, by = "customer_id") %>% select(gender, price) %>% group_by(gender) %>% summarise(avg_price = mean(price)) %>% show_query()
```

### Exercise 15: Failing Case with str_glue()
Try using `stringr::str_glue()` before `collect()`.

```{r}
# Your code here
customers_tbl %>% mutate(name = str_glue("{first_name} {last_name}")) %>% collect()

# this fails because there is no translation for str_glue in SQL.
```

### Exercise 16: Correct Use of `str_glue()`
Use `collect()` then use `str_glue()` safely.

```{r}
# Your code here
customers_tbl %>% collect() %>% mutate(name = stringr::str_glue("{first_name} {last_name}"))
```

## Theme 3: Independent Tasks

### Project A: Customer Insights
1. Count male vs. female customers.
2. Calculate average age by gender.
3. List top 3 customers with most purchases.
4. Create `full_name` using `paste()`.
5. Label age groups: "young" (<20), "adult" (20–49), "senior" (50+).
6. Export final tibble to CSV.

Count male vs. female customers.

```{r}
customers_tbl %>% group_by(gender) %>% count()
```

Calculate average age by gender.

```{r}
customers_tbl %>% 
  group_by(gender) %>% 
  summarize(avg_age = mean(age)) %>% 
  collect()
```

List top 3 customers with most purchases.

```{r}
sales_tbl %>% 
  count(customer_id, name = "sales_count") %>% 
  left_join(customers_tbl, by='customer_id') %>%
  mutate(full_name = paste(first_name, last_name, sep=" ")) %>%
  select(full_name, sales_count) %>% 
  slice_max(order_by = sales_count, n = 3, with_ties = F)

```

Label age groups: "young" (<20), "adult" (20–49), "senior" (50+).

```{r}
customer <- customers_tbl %>% mutate( age_groups = case_when(
  age < 20 ~ "young",
  age < 49 ~ "adult",
  T ~ "senior"
)) %>% collect()

customer
```

Export final tibble to CSV.
```{r}
# Use write.csv or readr::write_csv()
write.csv(customer, file="customer_data.csv")
```

### Project B: Sales Analysis
1. Count number of sales.
2. Calculate average sale price.
3. Filter for `price > 20` and count them.
4. Find the day with the highest total sales.
5. Join with customers and analyze total sales per gender.
6. Create a column `is_big_spender` if total_sales > 50.
7. Export the result to a file.

Count number of sales.

```{r}
sales_tbl %>% count()
```

Calculate average sale price.

```{r}
sales_tbl %>% summarise(avg_sale_price = mean(price))
```

Filter for `price > 20` and count them.

```{r}
sales_tbl %>% filter(price > 20) %>% count()
```

Find the day with the highest total sales.

```{r}
sales_tbl %>% count(sale_date, name = "daily_sale_count") %>% slice_max(order_by = daily_sale_count, n = 1)
```

Join with customers and analyze total sales per gender.

```{r}
# customers_tbl %>% left_join(sales_tbl, by = "customer_id") %>% group_by(gender) %>% summarise(avg_sales = mean(price))

customers_tbl %>% left_join(sales_tbl, by = "customer_id") %>% count(gender, name = "total_sales")

```

Create a column `is_big_spender` if total_sales > 50.

```{r}
h <- customers_tbl %>% left_join(sales_tbl, by = "customer_id") %>% count(gender, name = "total_sales") %>% mutate(is_big_spender = total_sales > 50)
h
```

Export the result to a file.

```{r}
write.csv(h, file = "big_spender.csv")
```



### Project C: Full Pipeline Practice
Write a complete pipeline that:
- Loads both tables
- Joins them
- Filters customers over 25 years old
- Computes average sales per customer
- Labels customer segment by total spending
- Collects and returns a final tibble
- Saves the result

```{r}
shop_db <- dbConnect(SQLite(), "shop.db")
cust_tbl <- tbl(shop_db, "customers")
sales_tbl <- tbl(shop_db, "sales")

g <- cust_tbl %>% left_join(sales_tbl, by="customer_id") %>% filter(age > 25) %>% group_by(customer_id) %>% summarise(avg_sales = mean(price), total_spending = sum(price))

write.csv(g, file="pipeline.csv")
```

### Advanced Reflection
- How would this change if the database had millions of rows?
- Where would you place `collect()` to ensure optimal performance?
- Can you modularize your pipeline using functions?

```{r}
# Disconnect when done
dbDisconnect(shop_db)
```

---

**Reflection:** Where did you place `collect()` in your pipelines and why? What issues did you face when trying to use non-translatable functions before `collect()`?

**Tip:** Always push as much filtering and summarising before `collect()` to minimize memory use and speed up your workflow.

**Challenge:** Replace your SQL-based workflows with full `{dplyr}` pipelines and compare performance and readability.

**Extension:** Explore the use of `dbWriteTable()` to write a new table to your SQLite database from a tibble. Try round-tripping a small summary table into the DB and back.
