---
title: "Project 2"
author: "Chantelle Amoako-Atta"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import libraries
```{r libraries}
library(DBI)
library(RSQLite)
library(dplyr)
library(tibble)
library(ggplot2)
```

# Part 1 - Obtaining the Data

## Connecting to the database
```{r database connection}
survey <- dbConnect(SQLite(), "likert_survey/likert_survey.db")
```

## Checking tables in the database
```{r}
dbListTables(survey)
```
Load student and professor tables
```{r}
student <- tbl(survey, "STUDENTS")
prof <- tbl(survey, "PROFESSORS")
```
## Previeing the tables' contents

### students
```{r}
student
```

### professors
```{r}
prof
```


# Part 2 - Exploring the data

## Q1 - What is the total number of answers obtained from students?

Get columns of student table
```{r}
colnames(student)
```


```{r}
total_students = student %>% 
  mutate(total_answers = TotallyDisagree+Disagree+Agree+TotallyAgree) %>% collect()

print(total_students)

paste("The total number of answers obtained from students is:", sum(total_students$total_answers))
```
## Q2 - What is the total number of answers obtained from professors?

Get columns of professors table
```{r}
colnames(prof)
```


```{r}
total_profs = prof %>% 
  mutate(total_answers = TotallyDisagree+Disagree+Agree+TotallyAgree) %>% collect()

print(total_profs)

paste("The total number of answers obtained from professors is:", sum(total_profs$total_answers))

```

## Q3 - Does each question have the same number of answers?

```{r}
print(unique(total_students$total_answers))
hist(total_students$total_answers)
```
From the histogram above, we observe that while most questions have 244 answers, some have less, implying that each question does not have the same number of students answering it.

Similarly, for professors:

```{r}
print(unique(total_profs$total_answers))
hist(total_profs$total_answers)
```

The results show us that all questions do not have the same number of answers, while most have the same number.

We could look deeper to see the questions that have fewer answers between students and professors.

### For students
```{r}
max_answers <- max(total_students$total_answers)


min_answered_q_students <- total_students %>% filter(total_answers < max_answers) %>% 
  select(QuestionID, total_answers) %>% 
  arrange(desc(total_answers))

ggplot(min_answered_q_students, aes(QuestionID, total_answers, fill = total_answers))+geom_col()
```

### For professors

```{r}
max_answers <- max(total_profs$total_answers)


min_answered_q_profs <- total_profs %>% filter(total_answers < max_answers) %>% 
  select(QuestionID, total_answers) %>% 
  arrange(desc(total_answers))

ggplot(min_answered_q_profs, aes(QuestionID, total_answers, fill = total_answers))+geom_col()
```




